PEACE: {\underline P}arallel {\underline E}nvironment for {\underline A}ssembly and {\underline C}lustering of Gene {\underline E}xpression

### 2010-02-12 papers/NAR_web_server_2009/paper.tex
Update to manuscript

\begin{abstract}
  We present {\sc Peace}, a standalone tool for the high-throughput
  {\it ab initio} clustering of transcript fragment sequences produced
  by Next Generation or Sanger Sequencing technologies.  Installed and
  run through an easily downloaded GUI, {\sc Peace} can process large
  data sets and group the fragments by gene association, achieving a
  greater sensitivity (without sacrificing runtime) than the competing
  clustering tool {\sc WCD}.  Through the GUI the user can collect
  cluster statistics and examine specific clusters for more
  comprehensive study.  {\sc Peace}, used as a ``pre-assembly'' tool,
  enables any assembly tool to work with individual small clusters of
  closely related sequences, resulting in a signficant: increases in
  sensitivity, robustness to sequenceing error, and reduction in
  overall runtime.
\end{abstract}

@@

Understanding an organism's transcriptome, the set of (spliced)
transcripts expressed by genes of the organism, is a vital step in
understanding the full functional and organizational role of the
genome in the life cycle of any eukaryote.  Studying the transcriptome
has led to gene discovery, provided information on splice variants,  
and helped shed light on the biological processes both controlling and
dictated by the genome \cite{Nagaraj07}.  But to access to those
transcripts, we must deal with the fragmented data produced by
both Next Generation and traditional Sanger sequencing technology.

Until recently, access to a transcriptome sequence was primarily
through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
sequences derived from transcribed mRNAs and sequenced by Sanger
Sequencing technology.  More recently, Next Generation Sequencing
technology (NGS) is rapidly replacing Sanger Sequencing, allowing for
much more comprehensive coverage of the transcriptome.  For example,
ESTs in GenBank dbEST are increasingly the product of NGS technologies
such as 454 pyrosequencing, which enables the sequencing of novel and
rare transcripts at a considerably higher rate of coverage than did
Sanger Sequencing \cite{Cheung2006,Emrich2007}.  From a computational
perspective, this is a mixed blessing: while NGS provides immense
quantities of new information, it also provides immensely larger data
sets, requiring fast computational techniques if we are to process
them.

Given a set of transcript fragments sampled from across the genome, a
necessary first step in the analysis is that of clustering: separating
the fragments according to the transcript from which they were
derived.  Frequently performed implicitly by assembly tools,
clustering the data as a ``pre-assembly'' step has a number of
advantages.  Most significant among these: performing this step will
allow the application of the assembly tool to individual clusters --
saving significant amounts of time due to the reduced input size
\cite{Hazelhurst08a}.  But clustering is a computationally challenging
problem.  Even with the smaller number of ESTs produced using Sanger
sequencing, the runtime and memory requirements to cluster based on
pair-wise sequence alignments make such an approach infeasible.  The
much larger data-set size produced by NGS 
technologies exacerbates this problem.

Here we present {\sc Peace}, a tool for the {\it ab initio} clustering
of transcript fragments by gene, applicable to both NGS and
traditional Sanger Sequencing technologies.  Available through the
\href{http://www.peace-tools.org}{www.peace-tools.org} website, the
{\sc Peace} GUI allows the user to both easily install (locally or remotely)
and run the computational tool, as well as enabling transparent
parallel processing and providing various tools for result analysis.
Using a {\it minimum spanning tree} (MST) based clustering strategy
\cite{Jain99,Wan08} and the $d^2$ sequence distance function
\cite{Hide94}, {\sc Peace} generates the MSTs using a tailored version
of Prim's algorithm \cite{Prim57}.  The result is an easy-to-use and
clustering tool that , without any sacrifice in runtime, is
considerably more sensitive and robust to sequencing errors than other
clusters tools in the literature \cite{Burke99,Slater00,
  Huang99,Parkinson02,Kalyanaraman03,Malde03,Ptitsyn05,Hazelhurst08a, Picardi09}.

@@

The {\sc Peace} GUI, available from the {\sc Peace} website
(\href{http://www.peace-tools.org}{www.peace-tools.org}), handles the
details of both installation and job management and provides tools for
the initial analysis of the results.  By downloading and executing the
{\tt peace.jar} GUI Java program on a Linux, Windows or OS-X machine
running Java 1.6+, the user is presented with the {\sc Peace}
management tool.  Through this the user can install the computational
tool and perform a clustering of a data file in the FASTA format, view an initial analysis
of the clusters and produce files containing subsets of the clusters
for assembly by a tool such as {\sc Cap3} \cite{Huang99}.  A typical
(first) use of {\sc Peace} would generally require the the following
steps.

\noindent {\bf Tool Installation:} Installing the GUI requires only
that the Java file be downloaded from the website.  To install the
{\sc Peace} tool onto a local or remote machine, the user will select
from within the GUI the appropriate menu tab (Figure~1(a)), which
then starts the install wizard that will prompt for the appropriate
information.  Figure~1(b) illustrates the request for server
information; the user has chosen to install the {\sc Peace}
computational tool on a remote machine and is providing the necessary
connection information.  Server information is persistent between GUI
sessions, giving the user access to {\sc Peace} on the target machine
as needed.

\noindent {\bf Job Processing:} After importing the target sequence file into
the GUI, the user can start a new job and follow the wizard menus.
Figure~1(c) illustrates the process of specifying the number of
processors available (if running on a machine supporting the OpenMPI
protocol -- which will be determined during job installation).  Once
executed, the GUI will manage the job thread, alert the user when the
job is completed (or when the user next runs the GUI after completion)
and copy the final results back to the local machine if necessary.

\noindent {\bf Result Analysis:} Once the resulting clusters have been
computed, the user has several options
for analysis:
\begin{itemize}
\item {\bf Export:} The user can export the contents of one or more
  clusters into a FASTA format file, obtaining a subset of the original target file
  containing the sequences corresponding to the selected clusters
  ready for processing by an assembly tool (e.g. {\sc Cap3} \cite{Huang99}).
\item {\bf View Clustering}: The user may view a list of clusters,
  expanding selected clusters to a list of all individual sequences (illustrated
  in Figure~1(d)).
\item {\bf Classified Summary Graph:} The user may view a distribution
  of cluster sizes.  Further, the user may set up a {\it classifier},
  associating certain patterns with specific colors.  These patterns
  will be matched against the fragment header information from the
  original FASTA file, allowing the overlay of a colored cluster size
  distributions.  For example, if the sequence names contain unique
  string patterns denoting different cDNA libraries, the classifier
  can help the user to determine and visualize the differential expression profiles
  of different libraries for a given cluster.  The method of setting up a these classifiers, and the
  resulting histogram, is illustrated in Figure~1(e).
\end{itemize}
Extensive documentation for the tool has been posted on the {\sc
  Peace} website, as well as links to several tutorial videos demonstrating
{\sc Peace} use and capabilities.

@@

We can model the fragment input as a weighted, undirected graph: the fragments
are represented as nodes, with $d^2$ sequence distances assigned to
the connecting edges as weights.  Conceptually, we want to remove each
edge exceeding a threshold score from the complete graph, and define
our partitions by the remaining connected components of the graph.  An
edge with a large weight connects fragments which are likely unrelated;
once such edges are removed, the components define a series of
overlaps.  Hence those fragments that can still be connected by some path
correspond to the same gene.  However, such an approach requires the
calculation of all edge weights.  Such a task is infeasible both in terms of
runtime and memory usage for the data set sizes we expect to process.

We approach the problem by generating a minimum spanning tree of the
described graph, then removing edges exceeding our threshold.  By
using Prim's algorithm we are able to calculate edge weights
on-the-fly (reducing memory requirements), and we can skip the
calculation of a majority of edge distances using the $u/v$ and $t/v$
filtering heuristics employed in {\sc wcd} \cite{Hazelhurst08a}.
These heuristics allow us to quickly dismiss many of the edges as too
large without the need to apply the full $d^2$ algorithm (see
Sections~A and B of the Supplementary Materials for more details).

@@

\begin{figure*}[b]
\centerline{
\includegraphics[scale=0.5]{pics.d/SeT1.pdf} 
}
\label{SeT1}
\caption{Comparisons of Sensitivity and Type 1 error, based on the
  average over 30 simulated fragment sets derived from 100 zebra fish genes
  (see Supplementary Materials, Section~\href{sim_results}, for more
  details).   Blue/Solid = {\sc Peace}, Green/Dash =
  {\sc wcd}, Black/Dot-Dash = {\sc Cap3}; vertical tics = 95\% confidence
  intervals on estimates.  Intervals are not presented for Type 1
  error due to the extremely small variance.}
\end{figure*}

\begin{figure*}[b]
\centerline{
\includegraphics[scale=0.35]{pics.d/singletons.pdf}
}
\label{singletons}
\caption{Average number of fragments flagged as singletons by each tool
  when run on the simulated sequences; correct answer in all cases is
  zero.  Blue/Solid = {\sc Peace}, Green/Dash = {\sc wcd},
  Black/Dot-Dash = {\sc Cap3}.}
\end{figure*}

@@

As a clustering tool based on sequence distance, {\sc Peace} faces
certain inherent limitations. For example, {\sc Peace} cannot handle
duplicate genes; like {\sc WCD}, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, naturally biological effects such as the trans-splicing of
transcripts, or effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters), post-transcriptional
poly(A)/(T) tails, and low-complexity repeats can cause similar
effects in these clustering tools.  The problems can be handled
through the application of the assembler, and the ability to apply any
assembler to small cluster (as opposed to the data set as a while)
results in a significant reduction in overall assembly time.

@@

Dr. Karro was funded under a PhRMA Foundation Informatics Research
Starters Grant while conducting this research.  We would also like to
acknowledge Iddo Friedberg, David Woods, Jens Mueller and David
Scoville at Miami University for their help with this project.

>>>>

\begin{abstract}
  We present {\sc Peace}, a standalone tool for the high-throughput
  {\it ab initio} clustering of transcript fragment sequences produced
  by Next Generation or Sanger Sequencing technologies.  Installed and
  run through an easily downloaded GUI, {\sc Peace} can process large
  data sets and group the fragments by gene association, achieving a
  greater sensitivity (without sacrificing runtime) than the competing
  clustering tools.  Through the GUI the user can collect
  cluster statistics and examine specific clusters for more
  comprehensive study.  {\sc Peace}, used as a ``pre-assembly'' tool,
  enables any assembly tool to work with individual small clusters of
  closely related sequences, resulting in signficant increases in
  sensitivit and robustness to sequenceing error, as well as a reduction in
  overall runtime.
\end{abstract}

@@

Understanding an organism's transcriptome, the set of (spliced)
transcripts expressed by genes of the organism, is a vital step in
understanding the full functional and organizational role of the
genome in the life cycle of any eukaryote.  Studying the transcriptome
has led to gene discovery, provided information on splice variants,  
and helped shed light on the biological processes both controlling and
dictated by the genome \cite{Nagaraj07}.  However, to access to those
transcripts, we must deal with the fragmented data produced by
both Next Generation and traditional Sanger sequencing technology.

Until recently, access to a transcriptome sequence was primarily
through the use of Expressed Sequence Tags (ESTs), single-pass cDNA
sequences derived from transcribed mRNAs and sequenced by Sanger
Sequencing technology.  More recently, Next Generation Sequencing
(NGS) technology has begun to rapidly replace Sanger Sequencing,
allowing for more comprehensive coverage of the transcriptome.  For
example, ESTs now being added to the GenBank dbEST are increasingly
the product of NGS technologies such as 454 pyrosequencing, which
enables the sequencing of novel and rare transcripts at a considerably
higher rate of coverage than Sanger Sequencing
\cite{Cheung2006,Emrich2007}.  From a computational perspective, this
is a mixed blessing: while NGS provides immense quantities of new
information, it also provides immensely larger data sets, requiring
faster computational techniques if we are to process them.

Given a set of transcript fragments sampled from across the genome, a
necessary first step in the analysis is that of clustering: separating
the fragments according to the transcript from which they were
derived.  Frequently performed implicitly by assembly tools,
clustering the data as a ``pre-assembly'' step has a number of
advantages.  Most significantly: performing this step will
allow the application of the assembly tool to individual clusters --
saving significant amounts of time due to the reduced input size
\cite{Hazelhurst08a}.  But clustering is a computationally challenging
problem.  Even with the smaller number of ESTs produced using Sanger
Sequencing, the runtime and memory requirements to cluster based on
pair-wise sequence alignments make such an approach infeasible in practice.  The
much larger data-set size produced by NGS 
technologies exacerbates this problem.

Here we present {\sc Peace}, a tool for the {\it ab initio} clustering
of transcript fragments by gene, applicable to both NGS and
traditional Sanger Sequencing technologies.  Available through the
\href{http://www.peace-tools.org}{www.peace-tools.org} website, the
{\sc Peace} GUI allows the user to both easily install (locally or remotely)
and run the computational tool, as well as enabling transparent
parallel processing and providing various tools for result analysis.
Using a {\it minimum spanning tree} (MST) based clustering strategy
\cite{Jain99,Wan08} and the $d^2$ sequence distance function
\cite{Hide94}, {\sc Peace} generates the MSTs using a tailored version
of Prim's algorithm \cite{Prim57}.  The result is an easy-to-use and
clustering tool that, without any sacrifice in runtime, is
considerably more sensitive and robust to sequencing errors than other
clusters tools in the literature \cite{Burke99,Slater00,
  Huang99,Parkinson02,Kalyanaraman03,Malde03,Ptitsyn05,Hazelhurst08a, Picardi09}.

@@

The {\sc Peace} GUI can be launched by downloading and executing the
Jar file available on the {\sc Peace} wesite
(\href{http://www.peace-tools.org}{www.peace-tools.org}) to any
machine running a standard Java Virtual Machine (JVM).  Once running,
the user can employ the GUI to install the computational tool and perform
a clustering of a data file in the FASTA format, view an initial
analysis of the clusters, and produce files containing subsets of the
clusters for assembly by a tool such as {\sc Cap3} \cite{Huang99}.
A typical (first) use of {\sc Peace} generally
requires the the following steps.

\noindent {\bf Tool Installation:} (First use only.)  To install the
{\sc Peace} clustering onto a local or remote machine, the
user selects from within the GUI the appropriate menu tab
(Figure~1(a)), which then starts an install wizard that will prompt
for the appropriate information.  Figure~1(b) illustrates the request
for server information; the user has chosen to install the {\sc Peace}
computational tool on a remote machine and is providing the necessary
connection information.  Server information is persistent between GUI
sessions, giving the user access to {\sc Peace} on the target machine
as needed.

\noindent {\bf Job Processing:} After importing the target sequence file into
the GUI, the user starts a new job by following the wizard menus.
Figure~1(c) illustrates the process of specifying the number of
processors available (if running on a machine supporting the OpenMPI
protocol -- which will be determined during job installation).  Once
executed, the GUI will manage the job thread, alert the user when the
job is completed (or when the user next runs the GUI after completion),
and copy the final results back to the local machine if necessary.

\noindent {\bf Result Analysis:} Once the resulting clusters have been
computed, the user has several options for analysis:
\begin{itemize}
\item {\bf Export:} The user can export the contents of one or more
  clusters into a FASTA format file, obtaining a subset of the original target file
  containing the sequences corresponding to the selected clusters
  ready for processing by an assembly tool (e.g. {\sc Cap3} \cite{Huang99}).
\item {\bf View Clustering}: The user may view a list of clusters,
  expanding selected clusters to a list of all individual sequences (illustrated
  in Figure~1(d)).
\item {\bf Classified Summary Graph:} The user may view a distribution
  of cluster sizes.  Further, the user may set up a {\it classifier},
  associating certain patterns with specific colors.  These patterns
  will be matched against the fragment header information from the
  original FASTA file, allowing the overlay of a colored cluster size
  distributions.  For example, if the sequence names contain unique
  string patterns denoting different cDNA libraries, the classifier
  can help the user to determine and visualize the differential expression profiles
  of different libraries for a given cluster.  The method of setting up these classifiers, and the
  resulting histogram, is illustrated in Figure~1(e).
\end{itemize}
Extensive documentation for the tool has been posted on the {\sc
  Peace} website, as well as links to several tutorial videos demonstrating
{\sc Peace} use and capabilities.

@@

We can model the fragment input as a weighted, undirected graph: the fragments
are represented as nodes, with $d^2$ sequence distances assigned to
the connecting edges as weights.  Conceptually, we want to remove each
edge exceeding a threshold score from the complete graph and define
our partitions by the remaining connected components of the graph.  An
edge with a large weight connects fragments which are likely unrelated;
once such edges are removed the components define a series of
overlaps.  Those fragments that can still be connected by some path
correspond to the same gene.  However, such an approach requires the
calculation of all edge weights.  Such a task is infeasible both in terms of
runtime and memory usage for the data set sizes we expect to process.

We approach the problem by generating a minimum spanning tree of the
described graph, then removing edges exceeding our threshold.  By
using Prim's algorithm we are able to calculate edge weights
on-the-fly (reducing memory requirements) and we can skip the
calculation of a majority of edge distances using the $u/v$ and $t/v$
filtering heuristics employed in {\sc wcd} \cite{Hazelhurst08a}.
These heuristics allow us to quickly dismiss many of the edges as too
large without the need to apply the full $d^2$ algorithm (see
Sections~A and B of the Supplementary Materials for more details).

@@

\begin{figure*}[h]
\centerline{
\begin{minipage}[b]{2in}
\includegraphics[scale=0.5]{pics.d/SeT1.pdf} 
\end{minipage}{2in}
}
\label{SeT1}
\caption{Comparisons of Sensitivity and Type 1 error, based on the
  average over 30 simulated fragment sets derived from 100 zebra fish genes
  (see Supplementary Materials, Section~\href{sim_results}, for more
  details).   Blue/Solid = {\sc Peace}, Green/Dash =
  {\sc wcd}, Black/Dot-Dash = {\sc Cap3}; vertical tics = 95\% confidence
  intervals on estimates.  Intervals are not presented for Type 1
  error due to the extremely small variance.}
\end{figure*}

\begin{figure*}[b]
\begin{minipage}{2in}
\centerline{
\includegraphics[scale=0.35]{pics.d/singletons.pdf}
}
\end{minipage}
\label{singletons}
\caption{Average number of fragments flagged as singletons by each tool
  when run on the simulated sequences; correct answer in all cases is
  zero.  Blue/Solid = {\sc Peace}, Green/Dash = {\sc wcd},
  Black/Dot-Dash = {\sc Cap3}.}
\end{figure*}

@@

As a clustering tool based on sequence distance, {\sc Peace} faces
certain inherent limitations. For example, {\sc Peace} cannot handle
duplicate genes; like {\sc WCD}, it is unable to separate clusters
corresponding to genes with a greater than 88\% similarity.
Similarly, natural biological effects such as the trans-splicing of
transcripts, or effects from poorly cleaned transcript data (e.g. the
failure to remove sequencing adapters), post-transcriptional
poly(A)/(T) tails, and low-complexity repeats can cause similar
effects in these clustering tools.  The problems can be handled
through the application of the assembler, and the ability to apply any
assembler to small cluster (as opposed to the data set as a whole)
results in a significant reduction in overall assembly time.

@@

Dr. Karro was funded under a PhRMA Foundation Informatics Research
Starters Grant while conducting this research.  We would also like to
acknowledge Iddo Friedberg, David Woods, Elizabeth Bikun, Jens Mueller and David
Scoville at Miami University for their help with this project.
